\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-07-03 submission/manuscript.tex
update for meeting

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, we
commonly see ML practices such as using ML pipelines where there is no
seperate held-out test dataset to evaluate model performance or
reporting few or only the best outcomes of cross-validation. Even when
there are seperate testing sets to evaluate model performance, we
observe large differences between cross-validation and testing
performances that indicate overfitting as well as large confidence
intervals for testing performances (4, 13--19). Moreover, there is a
lack of discussion on why a particular ML model is utilized. Recently,
there is a trend towards using more complex ML models such as random
forest, extreme gradient boosting and neural networks without a
discussion on if and how much model interpretibility is necessary for
the study (11, 20--22). Black box machine learning models require
posthoc explanations to determine the feature importances in making a
prediction. These explanations can be misleading and at times unreliable
when making high-stake decisions about someone's health (23). The lack
of transparency on model selection and interpretation as well as flawed
modeling methods negatively impact model validity and reproducibiliyu.
We need to strive toward better machine learning practices by (1)
implementing rigourous machine learning pipelines and (2) selecting ML
models that reflect the goal of the study as it will inform our
expectations of model accuracy, complexity, interpretibility and
computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and the same ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. CRC is a type of cancer which the human-associated human
microbiome is hypothesized to directly contribute to its development
therefore fecal 16S rRNA gene sequences can be used to detect CRC. We
built ML models using fecal 16S rRNA gene sequences to predict healthy
patients versus patients with colorectal lesions that were identified by
colonoscopy as screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate predictive performance and generalizability of each ML
model. We used the area under the receiver operating characteristic
curve (AUROC) as the predictive performance metric. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite its simplicity, the L2-regularized
logistic regression was second best in predictive performance. In terms
of computational efficiency, L1 SVM with linear kernel trained the
fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
(155.104 hours, std ± 0.959). We also found that depending on how the
data is split to create a held-out test set, the AUROC values of a ML
model can vary up to 0.295 which highlighted the importance of
performing many randomized modeling runs. This study established
standards for microbiome-associated ML models and underscored the
importance of model selection.

@@

We used a cohort of 490 patients with 261 cases of SRN. For each
patient, we had 6920 features (fecal bacterial abundances) and a
two-class label that defines their colonic health (SRN or normal). All
the cases were independently labeled through colonoscopies. We
established modeling pipelines for a binary prediction task with
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and extreme gradient boosted decision tree (XGBoost) to
emphasize the differences in model accuracy, complexity,
interpretibility and computational efficiency due to model selection.

@@

We established a ML pipeline where we train and validate each of the
seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection (i.e.~model selection), and the test set was
used for evaluation purposes. Validation of hyperparameter selection was
performed using repeated five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over the 100 randomizations and selected the best performing
hyper-parameter setting in terms of AUROC to train the full
training/validation dataset {[}Figures S1 and S2{]}. We then used the
held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model performance {[}Figure 1{]}.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel and L2 logistic regression using the feature
weights of the trained models. We ranked the absolute weights of all the
OTUs for each data-split. We calculated the median ranks of these
features over the 100 data-splits. In the three linear models, OTUs that
had the largest median ranks and drove the detection of SRNs belong to
family \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609),
genera \emph{Gamella} (OTU00426) and genera \emph{Peptostreptococcus}
(OTU00367) {[}Figure 3{]}. Some of the OTUs with the highest ranks were
shared among the linear models. We explained the feature importances in
non-linear models using permutation importance on the held-out test data
where we randomly permuted non-correlated features individually and
groups of correlated features together (see methods) to calculate the
effect of permuted OTUs or group of OTUs on testing AUROC. The top 5
OTUs with the largest negative impact on testing AUROC overlapped in
tree-based models {[}Figure 4{]}. Specifically, permuting
\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
predictive performances the most for all tree-based methods {[}Figure
S3{]}. Decision tree, random forest and XGBoost models' predictive
performance dropped from 0.6 base testing AUROC median to 0.52, from
0.69 to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.
Decision tree was more susceptible to The negative predictive impact of
\emph{Peptostreptococcus} in the decision tree model was followed by a
\emph{Lachnospiraceae} species (OTU00058) (0.6 base testing AUROC median
to 0.58) {[}Figure 4B{]}.

@@

As the complexity of a ML model increased {[}Table S1{]}, its training
times increased as well {[}Figure 5{]}. Additionally, the more
hyperparameters a model had {[}Table S1{]}, the longer it took to train
{[}Figure 5{]}. Linear models trained faster than non-linear models. L2
logistic Rregression and L1 and L2 SVM with linear kernel had training
times of 0.2 hours, (std ± 0.03), 0.2 hours, (std ± 0.02), and 0.2
hours, (std ± 0.02), respectively. Whereas, SVM with radial basis
function kernel, decision tree, random forest and xgboost had training
times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ± 8.8), 83.2 hours,
(std ± 11.3) and 155.1 hours, (std ± 1), respectively {[}Figure 5{]}.

@@

In this study we established a robust ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We used a held-out
test set to illustrate the difference between cross-validation and
testing AUROC values. When the differences between cross-validation and
test performance is low, this suggets the models will perform similar
with similar data. In all seven models, the differences in
cross-validation and testing AUROC values did not exceed 0.021 which
suggests that these models will be able to test similar new data.

In our ML pipeline we performed the 80\%-20\% random datasplit 100
times. The repetition of the initial data-split to create a held-out
test set is a crucial step in the ML pipeline to develop generalizable
ML models and to report reliable performance metrics for a ML model.
Depending on the data-split, there is the chance of being overoptimistic
about the predictive performance of a model. In our study, we showed
that there are differences in AUROC values between different random
data-splits in each of the models we tested. Our results showed that the
testing AUROC values varied 0.23 on average, depending on the
data-split.

Our results also showed that we should choose to use ML models based on
the goal of the study and our expectations of predictive performance,
interpretibility and computational burden. In terms of predictive
performance, all models except a decision tree and L1-regularized SVM
with linear kernel behaved similarly. Random forest model had testing
AUROC values statistically significantly larger than others. However the
second best model was L2-regularized logistic regression following
random forest with a median AUROC difference of only 0.015. The decision
tree model performed poorly because it had no regularization.
L1-regularized SVM with linear kernel also performed poorly possibly
because it was more sensitive to hyper-parameter tuning. In terms of
interpretation, random forest is a more complex ML model and it can only
be explained using methods such as permutation importance. On the other
hand, L2-regularized logistic regression is easy to interpret
(i.e.~regression coefficients of the trained model). If the goal of the
study is to learn the ecology behind a disease and to identify
biomarkers, then easily interpretable models could be a better fit. It
is also important to consider the computational burden of developing ML
models. Random forest model was slow to train with upto \texttt{r} hours
whereas L2-regularized logistic regression trained in \texttt{r}
minutes.

Our study underscored the importance of performing a full grid search
for hyperparameter settings when building a ML model. We first performed
a granular grid search and then a full grid search on the limited range
of possible hyperparameter settings. In the example of L1-regularized
SVM with linear kernel, the model showed variability between different
regularization coefficients (C).

\paragraph{Consideration of possible
weaknesses}\label{consideration-of-possible-weaknesses}

\begin{itemize}
\item
  What happens with imbalanced data The metric you use as the
  discriminative performance becomes important. Don't use accuracy.
  AUROC is good and random at 0.5 always.
\item
  What happens with smaller datasets Generalization performance depends
  on size. The more complex the model, more data you will need. Use
  linear models if you have little data.
\item
  Does the code work on any microbiome data? So if yu have a new sample,
  do you have the same OTUs? NO
\item
  Why did we use only one dataset?
\item
  Why not validate on other data instead of held-out testing set?
  Becuase we are comparing several methods on 1 dataset.
\item
  What will the future data look like?
\end{itemize}

\paragraph{Relationship of results to previous literature and broader
implications of having answered research
question}\label{relationship-of-results-to-previous-literature-and-broader-implications-of-having-answered-research-question}

\paragraph{Prospects for future
progress}\label{prospects-for-future-progress}

Subsample dataset and refit see how method behaves for 7 methods.
(500-\textgreater{}250-\textgreater{}50)

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
weights. The means weights and coefficients of the most important five
OTUs for each model are shown here with the standard deviation over 100
data-splits. Similar OTUs had the largest impact on the predictive
performance of L2 logistic regression and L2 SVM with linear kernel.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. For all the tree-based models,
a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
predictive performance of the model. Abbreviations: SVM, support vector
machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
Operational Taxonomic Unit. \newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The traintimes for training and testing
of each data-split showed the differences in computational efficieny of
the seven models. The median traintime in hours was the highest for
XGBoost and shortest for L2 logistic regression. The boxplot shows
quartiles at the box ends and the statistical median as the horizontal
line in the box. The whiskers show the farthest points that are not
outliers. Outliers are data points that are not within 3/2 times the
interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} Abbreviations: \newpage
\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} Abbreviations: \newpage

>>>>

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented {[}(12); dadkhah\_gut\_2019{]}. More notably, we
commonly see ML practices such as using ML pipelines where there is no
seperate held-out test dataset to evaluate model performance or
reporting few or only the best outcomes of cross-validation. Even when
there are seperate testing sets to evaluate model performance, there are
large differences between cross-validation and testing performances that
indicate overfitting as well as large confidence intervals for testing
performances (4, 13--19). Moreover, there is a lack of discussion on why
a particular ML model is utilized. Recently, there is a trend towards
using more complex ML models such as random forest, extreme gradient
boosting and neural networks without a discussion on if and how much
model interpretibility is necessary for the study (11, 20--22). Black
box machine learning models require posthoc explanations to determine
the feature importances in making a prediction. These explanations can
be misleading and at times unreliable when making high-stake decisions
about someone's health (23). The lack of transparency on model selection
and interpretation as well as flawed modeling methods negatively impact
model validity and reproducibiliy. We need to strive toward better
machine learning practices by (1) implementing rigourous machine
learning pipelines and (2) selecting ML models that reflect the goal of
the study as it will inform our expectations of model accuracy,
complexity, interpretibility and computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and the same ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. CRC is a type of cancer which the human-associated human
microbiome is hypothesized to directly contribute to its development and
fecal 16S rRNA gene sequences have been used to detect CRC. We built ML
models using fecal 16S rRNA gene sequences to predict healthy patients
versus patients with colorectal lesions that were identified by
colonoscopy as screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate predictive performance and generalizability of each ML
model. We used the area under the receiver operating characteristic
curve (AUROC) as the predictive performance metric. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite its simplicity, the L2-regularized
logistic regression was second best in predictive performance. In terms
of computational efficiency, L1 SVM with linear kernel trained the
fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
(155.104 hours, std ± 0.959). We also found that depending on how the
data is split to create a held-out test set, the AUROC values of a ML
model can vary up to 0.295 which highlighted the importance of
performing many randomized modeling runs. This study established
standards for microbiome-associated ML models and underscored the
importance of model selection.

@@

We used a cohort of 490 patients with 261 cases of SRN. For each
patient, we had 6920 features (fecal bacterial abundances) and a
two-class label that defines their colonic health (having colorectal
lesions that are identified as SRN or normal). All the cases were
independently labeled through colonoscopies. We established modeling
pipelines for a binary prediction task with L2-regularized logistic
regression, L1 and L2 support vector machines (SVM) with linear and
radial basis function kernels, a decision tree, random forest and
extreme gradient boosted decision tree (XGBoost) to emphasize the
differences in model accuracy, complexity, interpretibility and
computational efficiency due to model selection.

@@

We established a ML pipeline where we train and validate each of the
seven models {[}Figure 1{]}. We randomly split the data into
training/validation and test sets so that the training/validation set
consisted of 80\% of the full dataset while the test set was composed of
the remaining data {[}Figure 1{]}. Since the cases are not uniformly
represented in the data, the initial data-split was stratified to
maintain the overall label distribution in both the training/validation
and test sets. Training/validation set consisted of 393 patients (209
SRN), while the test set was composed of 97 patients (52 SRN). The
training/validation data was used for training purposes and validation
of hyperparameter selection (i.e.~model selection), and the test set was
used for evaluation purposes. Validation of hyperparameter selection was
performed using repeated five-fold cross-validation on the
training/validation set {[}Figure 1{]}. Similar to the initial
data-split, five-fold cross-validation was also stratified to maintain
the overall label distribution on the training and validation sets. We
validated the cross-validation performances of each hyperparameter
setting over 100 randomizations and selected the best performing
hyper-parameter setting in terms of AUROC metric and trained the full
training/validation dataset {[}Figures S1 and S2{]}. We then used the
held-out test set to evaluate the prediction performance of each ML
model. The data-split, hyperparameter selection, training and testing
steps were repeated 100 times to get a reliable and robust reading of
model performance {[}Figure 1{]}.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel and L2 logistic regression using the feature
weights of the trained models. We ranked the absolute weights of all the
OTUs for each data-split. We calculated the median ranks of these
features over the 100 data-splits. In the three linear models, OTUs that
had the largest median ranks and drove the detection of SRNs belong to
families \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} (OTU01239,
OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822, OTU00609),
genera \emph{Gamella} (OTU00426) and genera \emph{Peptostreptococcus}
(OTU00367) {[}Figure 3{]}. Some of the OTUs with the highest ranks were
shared among the linear models. We explained the feature importances in
non-linear models using permutation importance on the held-out test data
where we randomly permuted non-correlated features individually and
groups of correlated features together (see methods) to calculate the
effect of permuted OTUs or group of OTUs on testing AUROC. The top 5
OTUs with the largest negative impact on testing AUROC overlapped in
tree-based models {[}Figure 4{]}. Specifically, permuting
\emph{Peptostreptococcus} (OTU00367) abundances randomly, dropped the
predictive performances the most in all tree-based methods {[}Figure
S3{]}. Decision tree, random forest and XGBoost models' predictive
performance dropped from 0.6 base testing AUROC median to 0.52, from
0.69 to 0.68 and from 0.68 to 0.65, respectively {[}Figure 4{]}.

@@

As the complexity of a ML model and the number of tuned hyperparameter
settings increased {[}Table S1{]}, its training times increased as well
{[}Figure 5{]}. Linear models trained faster than non-linear models. L1
and L2 SVM with linear kernel and L2 logistic regression had training
times of 0.2 hours, (std ± 0.03), 0.2 hours, (std ± 0.02), and 0.2
hours, (std ± 0.02), respectively. Whereas, a decision tree, SVM with
radial basis function kernel, random forest and xgboost had training
times of 4.4 hours, (std ± 0.3), 59.6 hours, (std ± 8.8), 83.2 hours,
(std ± 11.3) and 155.1 hours, (std ± 1), respectively {[}Figure 5{]}.

@@

In this study we established a rigorous ML pipeline to use 16S rRNA
sequence counts to predict a binary health outcome. We set-up standards
for developing and evaluating ML models for microbiome data. We used a
held-out test set to illustrate the difference between cross-validation
and testing AUROC values. When the differences between cross-validation
and test performance is low, this suggets the models are not overfit and
that they will perform similar with similar data. In all seven models,
the differences in cross-validation and testing AUROC values did not
exceed 0.021 which suggests that these models will be able to test
similar new data. We performed the initial 80\%-20\% random datasplit
100 times in our ML pipeline. The repetition of the initial data-split
to create a held-out test set is a crucial step in the ML pipeline to
develop generalizable ML models and to report reliable performance
metrics for a ML model. Depending on the data-split, there is the chance
of being overoptimistic about the predictive performance of a model. In
our study, we showed that there was variability in AUROC values between
different random data-splits in each of the models we tested. Our
results showed that the testing AUROC values varied 0.23 on average,
depending on the data-split. We used AUROC metric instead of accuracy in
our study to evaluate predictive performance of the ML models. AUROC is
always random at the value 0.5 and is a robust metric when a dataset is
imbalanced. We performed a full grid search for hyperparameter settings
when building a ML model. It is not enough to use the default
hyperparameter settings when using previously developed ML packages in
programming languages like R, Python and Matlab. In the example of
L1-regularized SVM with linear kernel (Figure S1), the model showed
large variability between different regularization coefficients (C) and
was susceptible to performing poorly if the wrong regularization
coefficient was assigned to the model by default.

Our results showed that we should choose to use ML models based on the
goal of the study and our expectations of predictive performance,
interpretibility and computational burden. In terms of predictive
performance, random forest model had testing AUROC values statistically
significantly larger than others. However the second best model was
L2-regularized logistic regression following random forest with a median
AUROC difference of only 0.015. In terms of interpretation, random
forest was a more complex ML model and it could only be explained using
methods such as permutation importance. On the other hand,
L2-regularized logistic regression was easy to interpret
(i.e.~regression coefficients of the trained model). If the goal of a
study is to learn the ecology behind a disease and to identify microbial
biomarkers of a disease, then easily interpretable models would be
better suited. It is also important to consider the computational burden
of developing ML models. Random forest model was slow to train 83.2
hours whereas L2-regularized logistic regression trained in 12 minutes.
The dataset we used for our study had 490 samples. Another criteria when
choosing a ML model is the generalizability of it. The generalization
performance of ML models depends on size. The more complex the model,
the more data it will need. Microbiome studies that have small sample
sizes should use less complex models such as L2-regularized logistic
regression.

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
ranks. The feature ranks of the most important five OTUs in each
data-split are shown here. Similar OTUs had the largest impact on the
predictive performance of L2 logistic regression and L2 SVM with linear
kernel. Abbreviations: SVM, support vector machine; OTU, Operational
Taxonomic Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. For all the tree-based models,
a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
predictive performance of the model. Abbreviations: SVM, support vector
machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
Operational Taxonomic Unit. \newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The training times for training and
testing of each data-split showed the differences in computational
efficieny of the seven models. The median traintime in hours was the
highest for XGBoost and shortest for L2 logistic regression. The boxplot
shows quartiles at the box ends and the statistical median as the
horizontal line in the box. The whiskers show the farthest points that
are not outliers. Outliers are data points that are not within 3/2 times
the interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} Abbreviations: \newpage
\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} Abbreviations: \newpage
